Copulas and their potential for ecology: Appendices

### 2019-12-23 SupportingInformation.Rmd
minor corrections made while checking Shyas conversion to Word

\textbf{Community-level data, Cedar Creek.} Above-ground biomass [@ccdataple120] and percent cover data [@ccdatapce120] were 
obtained from biodiversity experiment E120 conducted at the Cedar Creek Ecosystem Science Reserve 
by Tilman *et al*. In 1994, 168 plots [@ccdataplotse120] (each 9 m $\times$ 9 m) of 39, 35, 29, 30 and 35 replicates,
were seeded with 1, 2, 4, 8, or 16 perennials, respectively, chosen randomly from a pool of 
18 species (4 species, each, of C4 grasses, C3 grasses, legumes, non-legume forbs; and 2 species of woody
plants). All plots received, in total, 10 $gm^{-2}$ of pure live seed in May 1994 and 5 $gm^{-2}$ in May, 1995 
with seed mass divided equally among species. Treatments were maintained by weeding manually 3-4 times a year 
and sometimes selective herbicides were used. We downloaded the data on July, 2018 from the Cedar Creek website \url{http://www.cedarcreek.umn.edu/research/experiments/e120} and 
analyzed those data for 6 years 
(1996-2000, 2007) for which both biomass and percent cover data were available. Plots were annually 
sampled by clipping 
different locations each year. For each available year, we analyzed above-ground biomass ($gm^{-2}$) 
and Shannon's biodiversity index (computed using percent cover data) averaged over the 
sampled strips/quadrats taken from each 
of 168 plots.

@@

knitr::kable(NA, "latex", booktabs = T, linesep = "", escape=F,
             caption.short = "Summary of principal notation",
      caption = "Summary of principal notation. Many notations are defined and used only within a small part of the paper. This table summarizes notation that is used across larger portions of the paper. Third coloumns to the section in which the notation was originally defined. cdf = cumulative distribution function. \\label{tab_summary_notation}")
```

@@

<!--embedding plot with sampling locations : soilCN, aphids, planktons-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=16 cm]{./Results/SoilCNmapUSA.pdf} 
\caption[Sampling locations for C and N data]{Locations of central pedons sampled 
for soil organic carbon and total soil nitrogen (Mg C or N per hectare of soil 
surface) at 5907 sites across the coterminous United States. \label{fig_SoilCNmapUSA}}
\end{center}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_BB8.pdf}
\caption[Example BB8 copulas]{Log-transformed pdfs for example BB8 copulas. K is Kendall correlation; $p_1$ and $p_2$
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, 
respectively. The parameter ranges for the family are $p_1 \in [1,\infty)$
and $p_2 \in (0,1]$, lower-tail dependence is $0$ and upper-tail
dependence is $2-2^{1/p_1}$ of $p_2=1$ and $0$ otherwise.\label{figped_BB8}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_SBB8.pdf}
\caption[Example survival BB8 copulas]{Log-transformed pdfs for example survival BB8 copulas. K is Kendall correlation; $p_1$ and $p_2$ 
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, 
respectively. The parameter ranges for the family are $p_1 \in [1,\infty)$
and $p_2 \in (0,1]$, upper-tail dependence is $0$ and lower-tail
dependence is $2-2^{1/p_1}$ of $p_2=1$ and $0$ otherwise.\label{figped_SBB8}}
\end{center}
\end{figure}

@@

\noindent We define $\Ps_{l_b,u_b}$, using the bounds
$u+v=2l_b$ and $u+v=2u_b$ defined in the main text, as follows. For a distance $h$, we define
$S(h)$ to be the number of points $(u_i,v_i)$ within the bounds 
and a distance less than $h$ from the line $v=u$, divided by the 
total number of points within the bounds. This function is
defined for $h$ going from $0$ to a distance $h_{\text{max}}$ which
is half the longer of the two segments obtained by intersecting the
bounds with the unit square. It is easy to see $S(0)=0$ and 
$S(h_{\text{max}})=1$. We define $S_i(h)$, also for $h$ going from $0$
to $h_{\text{max}}$, to be the area within the bounds and the unit square
and within distance $h$ of the line $v=u$, divided by the area
within the bounds and the unit square. This is the expected value of
$S(h)$ for independent data. We then define
\begin{equation}\label{eq.P}
   P_{l_b,u_b} = \int_0^{h_{\text{max}}} ( S(h) - S_i(h) ) dh.
\end{equation}
\noindent As for $\cor_{l_b,u_b}$, larger values of $\Ps_{l_b,u_b}$ 
indicate stronger positive association
between $u$ and $v$ in the region given by the bounds.
We use both statistics 
because $\cor_{l_{b_k},u_{b_k}}(u,v)$ is more familiar, but 
in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_testing/res_pt35_rankT.pdf}
\caption[Testing nonparametric statistics with 35 data points]{Results of tests of whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ reveal the known asymmetry of tail dependence of the Clayton and survival Clayton copulas, and the known symmetry of tail dependence of the normal and Frank copulas. p-values (indicated by the triangles) when less than $0.05$ (below the dashed line) mean there is significant tail asymmetry. See Appendix  \ref{Test_npa_stat} for details. $n=35$. \label{fig_stat_testing35}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_testing/res_pt1000_rankT.pdf}
\caption[Testing nonparametric statistics with 1000 data points]{Results of tests of whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ reveal the known asymmetry of tail dependence of the Clayton and survival Clayton copulas, and the known symmetry of tail dependence of the normal and Frank copulas. p-values (indicated by the triangles) when less than $0.05$ (below the dashed line) mean there is significant tail asymmetry. See Appendix  \ref{Test_npa_stat} for details. $n=1000$.\label{fig_stat_testing1000}}
\end{center}
\end{figure}

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible.
using the `iRho` function (or the `iTau` function) of the `copula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k^{th}$ smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is a surrogate dataset.

\noindent Our code for this algorithm is `ncsurrog` in the `BIVAN` repository 
(\url{https://github.com/sghosh89/BIVAN}).
The code for the Kendall version of the
algorithm throws an error if there are ties in any of the time series. This is
because the Kendall correlation in the very commonly used `cor` function in R 
is the Kendall tau-b coefficient in the event of ties, and index which 
differs in its definition from the commonly used tau-a that works when there are
no ties; and `tau` and `iTau` are intended for continuous copulas, which cannot 
produce data with ties. Our code also throws an error in the event that
the matrix generated in the second step is not positive semidefinite. Because
ties are common in our data, we only used the Spearman case in analyses using `ncsurrog`,
though, for expansibility, the `ncsurrog` code itself is written to handle 
the Kendall case as well when there are no ties.

@@

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-paremeter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distributions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N},k)}_{t,j}$ is our normal-copula population array (the three dimensions 
of the array are indexed, respectively, by $t$, $j$ and $k$, and $m^{(\text{C},k)}_{t,j}$ is our 
Clayton-copula population array, then,
for any given $j$ and $k$, the sets $\{m^{(\text{N},k)}_{t,j}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C},k)}_{t,j}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $j_1$, $j_2$ and $k$ with $j_1 \neq j_2$, the Spearman correlations $\cor_t(m^{(\text{C},k)}_{t,j_1},m^{(\text{C},k)}_{t,j_2})$
and $\cor_t(m^{(\text{N},k)}_{t,j_1},m^{(\text{N},k)}_{t,j_2})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

@@

- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent 
draws from $C$.
- For each $k$ and $i$, replace the $k^{th}$-smallest element of the time series $a_i(t)$, 
$t=1,\ldots,T$ by the $k^{th}$ smallest element of $x_i(t)$, $t=1,\ldots,T$. Call the 
result $y_i(t)$. This is the surrogate dataset.

>>>>

\textbf{Community-level data, Cedar Creek.} Above-ground biomass [@ccdataple120] and percent cover data [@ccdatapce120] were 
obtained from biodiversity experiment E120 conducted at the Cedar Creek Ecosystem Science Reserve 
by Tilman *et al*. In 1994, 168 plots [@ccdataplotse120] (each 9 m $\times$ 9 m) of 39, 35, 29, 30 and 35 replicates,
were seeded with 1, 2, 4, 8, or 16 perennials, respectively, chosen randomly from a pool of 
18 species (4 species, each, of C4 grasses, C3 grasses, legumes, non-legume forbs; and 2 species of woody
plants). All plots received, in total, 10 gm$^{-2}$ of pure live seed in May 1994 and 5 gm$^{-2}$ in May, 1995 
with seed mass divided equally among species. Treatments were maintained by weeding manually 3-4 times a year 
and sometimes selective herbicides were used. We downloaded the data on July, 2018 from the Cedar Creek website \url{http://www.cedarcreek.umn.edu/research/experiments/e120} and 
analyzed those data for 6 years 
(1996-2000, 2007) for which both biomass and percent cover data were available. Plots were annually 
sampled by clipping 
different locations each year. For each available year, we analyzed above-ground biomass (gm$^{-2}$) 
and Shannon's biodiversity index (computed using percent cover data) averaged over the 
sampled strips/quadrats taken from each 
of 168 plots.

@@

knitr::kable(NA, "latex", booktabs = T, linesep = "", escape=F,
             caption.short = "Summary of principal notation",
      caption = "Summary of principal notation. Many notations are defined and used only within a small part of the paper. This table summarizes notation that is used across larger portions of the paper. Third column refers to the section in which the notation was originally defined. cdf = cumulative distribution function. \\label{tab_summary_notation}")
```

@@

<!--embedding plot with sampling locations : soilCN, aphids, planktons-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=16 cm]{./Results/SoilCNmapUSA.pdf} 
\caption[Sampling locations for C and N data]{Locations of central pedons sampled 
for soil organic carbon and total soil nitrogen (megagrams C or N per hectare of soil 
surface) at 5907 sites across the coterminous United States. \label{fig_SoilCNmapUSA}}
\end{center}
\end{figure}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_BB8.pdf}
\caption[Example BB8 copulas]{Log-transformed pdfs for example BB8 copulas. K is Kendall correlation; $p_1$ and $p_2$
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, 
respectively. The parameter ranges for the family are $p_1 \in [1,\infty)$
and $p_2 \in (0,1]$, lower-tail dependence is $0$ and upper-tail
dependence is $2-2^{1/p_1}$ if $p_2=1$ and $0$ otherwise.\label{figped_BB8}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_SBB8.pdf}
\caption[Example survival BB8 copulas]{Log-transformed pdfs for example survival BB8 copulas. K is Kendall correlation; $p_1$ and $p_2$ 
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, 
respectively. The parameter ranges for the family are $p_1 \in [1,\infty)$
and $p_2 \in (0,1]$, upper-tail dependence is $0$ and lower-tail
dependence is $2-2^{1/p_1}$ if $p_2=1$ and $0$ otherwise.\label{figped_SBB8}}
\end{center}
\end{figure}

@@

\noindent We define $\Ps_{l_b,u_b}$, using the bounds
$u+v=2l_b$ and $u+v=2u_b$ defined in the main text, as follows. For a distance $h$, we define
$S(h)$ to be the number of points $(u_i,v_i)$ within the bounds 
and a distance less than $h$ from the line $v=u$, divided by the 
total number of points within the bounds. This function is
defined for $h$ going from $0$ to a distance $h_{\text{max}}$ which
is half the longer of the two segments obtained by intersecting the
bounds with the unit square. It is easy to see $S(0)=0$ and 
$S(h_{\text{max}})=1$. We define $S_i(h)$, also for $h$ going from $0$
to $h_{\text{max}}$, to be the area within the bounds and the unit square
and within distance $h$ of the line $v=u$, divided by the area
within the bounds and the unit square. This is the expected value of
$S(h)$ for independent data. We then define
\begin{equation}\label{eq.P}
   P_{l_b,u_b} = \int_0^{h_{\text{max}}} ( S(h) - S_i(h) ) dh.
\end{equation}
\noindent As for $\cor_{l_b,u_b}$, larger values of $\Ps_{l_b,u_b}$ 
indicate stronger positive association
between $u$ and $v$ in the region given by the bounds.
We use both statistics 
because $\cor_{l_{b},u_{b}}(u,v)$ is more familiar, but 
in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_testing/res_pt35_rankT.pdf}
\caption[Testing nonparametric statistics with 35 data points]{Results of tests of whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ reveal the known asymmetry of tail dependence of the Clayton and survival Clayton copulas, and the known symmetry of tail dependence of the normal and Frank copulas. p-values (indicated by the triangles), when less than $0.05$ (below the dashed line), mean there is significant tail asymmetry. See Appendix  \ref{Test_npa_stat} for details. $n=35$. \label{fig_stat_testing35}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/stat_results/stat_testing/res_pt1000_rankT.pdf}
\caption[Testing nonparametric statistics with 1000 data points]{Results of tests of whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ reveal the known asymmetry of tail dependence of the Clayton and survival Clayton copulas, and the known symmetry of tail dependence of the normal and Frank copulas. p-values (indicated by the triangles), when less than $0.05$ (below the dashed line), mean there is significant tail asymmetry. See Appendix  \ref{Test_npa_stat} for details. $n=1000$.\label{fig_stat_testing1000}}
\end{center}
\end{figure}

@@

- Given data $x_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$, let $\tau_{ij}$ be the Spearman (or Kendall) 
correlation of $x_i(t)$ and $x_j(t)$.
- Calculate the covariance matrix of a multivariate normal distribution with standard-normal
marginals and with pairwise Spearman (or Kendall) correlations equal to the $\tau_{ij}$. This is possible.
using the `iRho` function (or the `iTau` function) of the `copula` package
- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent draws
from the multivariate normal distribution of the previous step.
- For any $i,t$ for which the datum $x_i(t)$ was missing, replace $a_i(t)$ by `NA`.
- For each $k$ and $i$ for which $a_i(t)$ is not `NA`, replace the 
$k$th-smallest (non-`NA`) element of the time series 
$a_i(t)$, $t=1,\ldots,T$ by the $k^{\text{th}}$ smallest (non-`NA`) 
element of $x_i(t)$, $t=1,\ldots,T$.
Call the result $b_i(t)$. This is a surrogate dataset.

\noindent Our code for this algorithm is `ncsurrog` in the `BIVAN` repository 
(\url{https://github.com/sghosh89/BIVAN}).
The code for the Kendall version of the
algorithm throws an error if there are ties in any of the time series. This is
because the Kendall correlation in the very commonly used `cor` function in R 
is the Kendall tau-b coefficient in the event of ties, an index which 
differs in its definition from the commonly used tau-a that works when there are
no ties; and `tau` and `iTau` are intended for continuous copulas, which cannot 
produce data with ties. Our code also throws an error in the event that
the matrix generated in the second step is not positive semidefinite. Because
ties are common in our data, we only used the Spearman case in analyses using `ncsurrog`,
though, for expansibility, the `ncsurrog` code itself is written to handle 
the Kendall case as well when there are no ties.

@@

We then generated a series of arrays that were the same dimensions ($N \times n \times 1000$)
as the normal-copula population array, and that were statistically similar to 
the normal-copula population array in all ways except the new
arrays were constructed to have Clayton (respectively, Frank, survival Clayton) copula structure for 
the dependences between 
populations in different sampling locations within a simulation. We 
refer to the new arrays as our
*Clayton-copula*, *Frank-copula*, and *Survival Clayton-copula population arrays*, respectively. The Clayton-copula array was
constructed as follows; the others were constructed analogously. First, we constructed an
$n$-dimensional Clayton copula with parameter chosen so that the 
Spearman correlations between each pair of the $n$ components were the same as those
of the normal distribution constructed above. This was done using the `rho` and `iRho` functions 
in the `copula` package in R. Because the Clayton copula is a one-parameter family, this 
step relies on the fact that all pairs of locations were 
equally correlated with each other in the original multivariate normal distribution
constructed above. We then randomized each simulation in 
the normal-copula population array in a special way 
(see Appendix \ref{sect:copsurrognd}) so that the resulting 
surrogate/randomized dataset had the copula structure
of the Clayton copula just constructed. The randomization procedure worked 
by permuting time series of values in the normal-copula array, separately for each 
sampling location and simulation. So population marginal distributions for each location
and simulation were *exactly* the same as those in the normal-copula population array. 
Permutations were done in such a way
that resulting pairwise Spearman correlations were the same as the Clayton copula 
constructed above, and hence were the same as the normal-copula array, 
except for sampling variation. Thus, to summarize using mathematical notation,
if $m^{(\text{N},k)}_{t,j}$ is our normal-copula population array (the three dimensions 
of the array are indexed, respectively, by $t$, $j$ and $k$), and $m^{(\text{C},k)}_{t,j}$ is our 
Clayton-copula population array, then,
for any given $j$ and $k$, the sets $\{m^{(\text{N},k)}_{t,j}: t=1,\ldots,N\}$ and 
$\{m^{(\text{C},k)}_{t,j}: t=1,\ldots,N\}$ are exactly the same (when considered 
as unordered sets); 
and for any given $j_1$, $j_2$ and $k$ with $j_1 \neq j_2$, the Spearman correlations $\cor_t(m^{(\text{C},k)}_{t,j_1},m^{(\text{C},k)}_{t,j_2})$
and $\cor_t(m^{(\text{N},k)}_{t,j_1},m^{(\text{N},k)}_{t,j_2})$ are the same, to within 
sampling variation. The whole exercise was repeated preserving Kendall instead of 
Spearman correlations.

@@

- Generate data $a_i(t)$ for $i=1,\ldots,N$ and $t=1,\ldots,T$ by taking $T$ independent 
draws from $C$.
- For each $k$ and $i$, replace the $k^{\text{th}}$-smallest element of the time series $a_i(t)$, 
$t=1,\ldots,T$ by the $k^{\text{th}}$ smallest element of $x_i(t)$, $t=1,\ldots,T$. Call the 
result $y_i(t)$. This is the surrogate dataset.
